\begin{abstract}
    \small
    RNA sequencing is a popular assay for measuring transcriptomes and
    is useful for expression quantification, and sequence assembly of genes and transcripts.
    The vast amount of available raw RNA sequencing data, and its growth rate over the past decade
    brings us to an important computational challenge prior to any higher level analyses;
    The challenge of designing a memory-efficient data structure for indexing the underlying set of sequences
    that supports a time-efficient query. This challenge is not limited to only raw RNA sequences.
    By the continuing assembly of novel genomes and transcriptomes,
    specifically in metagenomics, having an efficient index over the list of assembled sequences is
    still an ongoing challenge. To be specific,
    there is still a high demand for an index providing a balance between memory and query time.
    We have developed new data structures (both reference-based and reference-free)
    to substantially reduce the memory used by these indices
    and still achieve similar query speed.
%    which can divide these data structures into two categories of reference-based and reference-free indices.

%    These are highly promising data structures for sequence search, sequence alignment, or sequence assembly
%    and are useful for transcriptomic, genomic, metagenomic, and microbiome analysis.


    We propose three separate indices, ``Pufferfish'', an index over a set of genomes or transcriptomes,
    ``Rainbowfish'' and ``Mantis'' which are indices for indexing a set of raw sequencing data sets.
%    one over the set of reference sequences used for alignment and mapping applications called Pufferfish,
%    one over a set of raw sequences designed to be accessed for assembly, variant detection and bubble calling
%    called Rainbowfish,
%    and a third one also being over a large set of sequences, but used in sequence search applications called Mantis.
%    All these data structures are based on the de Bruijn graph and different variations of it
%    such as the colored de Bruijn graph or the compacted de Bruijn graph.
    Specifically we designed a new, succinct representation for the color information
    of colored de Bruijn graphs (cdbg) in “Rainbowfish”.
    This representation can be used in de novo assembly and variant detection
    to keep information about the sample of origin of each k-mer when combining many samples.
    For example, our data structure allows building a cdbg
    on a metagenomic data set in just a few gigabytes
    while the space taken by other structures is hundreds of gigabytes,
    showing an order-of-magnitude improvement.
    Moreover, the query time for searching a k-mer and fetching the color information
    in our representation is almost the same as the existing data structures.

    Using a similar representation as Rainbowfish to store the list of samples each k-mer appears in
    , combined with the use of the counting quotient filter to find the index associated to each k-mer,
    we've developed Mantis,
    a space-frugal index over thousands of samples with a fast query time up to 108 times faster than
    the sequence search tools at the time.
    Mantis is also a cdbg representation which supports fast graph traversal and other topological analyses
    in addition to large-scale sequence-level searches. Later, we update the color information representation
    by adopting a hierarchical encoding that exploits correlations among color vectors.
    Our results show significant improvements in the overall size and
    scalability of representation of the color information.

    In Pufferfish, however, we've developed a new data structure for indexing %large-scale reference sequences, or
    large collections of reference sequences rather than raw sequence samples by building an index
    for the compacted, colored de Bruijn graph (ccdbg) over the reference sequences.
    We have developed both a sparse and dense indexing scheme which allows one
    to trade off index space for query speed
    (though queries always remain asymptotically optimal).
    In addition to indexing the k-mers, we store informations such as
    position and orientation of each k-mer in the reference set
    which makes it useful for purposes such as alignment and quantification.

%    In designing all the above methods, we make use of techniques
%    from succinct data structure design — building more complex operations atop bit vectors
%    and the rank and select operations and also minimum perfect hashing to help substantially reduce the memory
%    required for compacted de Bruijn graph indexing.

    For my thesis, I propose advancing the ideas explained earlier in two specific, but separate directions.
    For the reference-base index scheme, I focus on one of the applications of Pufferfish as the alignment tool
    used in estimating abundance of known genomes
    in metagenomic analysis. This work is along the lines of the tools
    such as Bracken and Karp and slightly different from
    metagenomic classification methods such as Kraken. In addition to that, I will continue working on
    the scalability limits of Mantis as a large-sequence-search index by investigating the potentials of
    merging two Mantises without the need to construct the color information in bit vector format and convert it
    to the hierarchical structure and by controlling the construction memory in a limited range
    with respect to the size of the input data.


    % de Bruijn graphs are nowadays an inseparable part of Next Generation Sequence analyses. With the fast growth in the amount of sequencing reads and variety of different genomes and transcriptomes the community is shifting toward using graphs to represent collections of references and map reads.
    % In this document, we present two succinct representations for two different variations of a de Bruijn graph, namely colored de Bruijn graph and compacted de Bruijn graph. For the former, we designed and developed rainbowfish, a succinct data structure to represent colors for an efficient \dbg representation and also theoretically proved why we call it succinct. This structure is useful in genome variant detection and genotyping. For the later, we developed a tool named pufferfish for indexing a compacted de Bruijn graph in two different schemes of dense and sparse. While being close to linear indexing methodologies regarding memory, pufferfish shows to have a similar \kmer lookup speed as other memory-consuming de Bruijn graph indexing schemes. The balance that pufferfish offers between memory and speed makes it suitable for large-scale indexing such as for collections of RNA-seq data or in metagenomic analysis.
\end{abstract} 