%%
%% Author: fatemeh
%% 2019-11-25
%%

\chapter{Reading List}
\label{chap:readingList}

List of 30 papers:
\subsubsection{Counting Kmers}
\begin{enumerate}
    \item Marçais, Guillaume, and Carl Kingsford. "A fast, lock-free approach for efficient parallel counting of occurrences of k-mers." Bioinformatics 27.6 (2011): 764-770.
    \item Mohamadi, Hamid, Hamza Khan, and Inanc Birol. "ntCard: a streaming algorithm for cardinality estimation in genomics data." Bioinformatics 33.9 (2017): 1324-1330.
%    \item Kokot, Marek, Maciej Długosz, and Sebastian Deorowicz. "KMC 3: counting and manipulating k-mer statistics." Bioinformatics 33.17 (2017): 2759-2761.
\end{enumerate}
\subsubsection{De bruijn graphs → definition, use cases, On the representation of dbgs}
\begin{enumerate}
    \item Rayan Chikhi, Antoine Limasset, Shaun Jackman, Jared T Simpson, and Paul Medvedev. On the representation of de bruijn graphs. In International conference on Research in computational molecular biology, pages 35–55. Springer, 2014.
    \item Bowe, Alexander, et al. "Succinct de Bruijn graphs." International Workshop on Algorithms in Bioinformatics. Springer, Berlin, Heidelberg, 2012.
\end{enumerate}
\subsubsection{Compacted de Bruijn graph, Direct Construction from Raw Read/Reference Sequences}
\begin{enumerate}
    \item Chikhi, Rayan, Antoine Limasset, and Paul Medvedev. "Compacting de Bruijn graphs from sequencing data quickly and in low memory." Bioinformatics 32.12 (2016): i201-i208.
    \item Minkin, Ilia, Son Pham, and Paul Medvedev. "TwoPaCo: An efficient algorithm to build the compacted de Bruijn graph from many complete genomes." Bioinformatics 33.24 (2016): 4024-4032.
\end{enumerate}
\subsubsection{Colored de Bruijn graph}
\begin{enumerate}
    \item Martin D. Muggli, Alexander Bowe, Noelle R. Noyes, Paul Morley, Keith Belk, Robert Raymond, Travis Gagie, Simon J. Puglisi, and Christina Boucher. Succinct Colored de Bruijn Graphs. Bioinformatics, 2017.
    \item Bahar Alipanahi, Martin D Muggli, Musa Jundi, Noelle Noyes, and Christina Boucher. Resistome SNP calling via read colored de Bruijn graphs. bioRxiv, page 156174, 2018.
    \item Bahar Alipanahi, Alan Kuhnle, and Christina Boucher. Recoloring the Colored de Bruijn Graph. In International Symposium on String Processing and Information Retrieval, pages 1–11. Springer, 2018.
\end{enumerate}
\subsubsection{Minimal Perfect Hash function}
\begin{enumerate}
    \item Limasset, Antoine, et al. "Fast and scalable minimal perfect hashing for massive key sets." arXiv preprint arXiv:1702.03154 (2017).
\end{enumerate}
\subsubsection{Approximate Membership Queries}
\begin{enumerate}
    \item Putze, Felix, Peter Sanders, and Johannes Singler. "Cache-, hash-, and space-efficient bloom filters." Journal of Experimental Algorithmics (JEA) 14 (2009): 4.
    \item Salikhov, Kamil, Gustavo Sacomoto, and Gregory Kucherov. "Using cascading Bloom filters to improve the memory usage for de Brujin graphs." Algorithms for Molecular Biology 9.1 (2014): 2.
    \item Pandey, Prashant, et al. "A general-purpose counting filter: Making every bit count." Proceedings of the 2017 ACM International Conference on Management of Data. ACM, 2017.
\end{enumerate}
\subsubsection{Large Short Sequence Database Search}
\begin{enumerate}
    \item Solomon, Brad, and Carl Kingsford. "Fast search of thousands of short-read sequencing experiments." Nature biotechnology 34.3 (2016): 300.
    \item Solomon, Brad, and Carl Kingsford. "Improved search of large transcriptomic sequencing databases using split sequence bloom trees." International Conference on Research in Computational Molecular Biology. Springer, Cham, 2017.
    \item Sun, Chen, et al. "Allsome sequence bloom trees." Journal of Computational Biology 25.5 (2018): 467-479.
    \item Harris, Robert S., and Paul Medvedev. "Improved representation of sequence Bloom trees." BioRxiv (2018): 501452.
    \item Bradley, Phelim, et al. "Ultrafast search of all deposited bacterial and viral genomic data." Nature biotechnology 37.2 (2019): 152.
    \item Yu, Ye, et al. "SeqOthello: querying RNA-seq experiments at scale." Genome biology 19.1 (2018): 167.
    \item Holley, Guillaume, Roland Wittler, and Jens Stoye. "Bloom Filter Trie: an alignment-free and reference-free data structure for pan-genome storage." Algorithms for Molecular Biology 11.1 (2016): 3.
\end{enumerate}
\subsubsection{Next Generation Sequence (short-read) Alignment}
\begin{enumerate}
    \item Langmead, Ben, and Steven L. Salzberg. "Fast gapped-read alignment with Bowtie 2." Nature methods 9.4 (2012): 357.
    \item Dobin, Alexander, et al. "STAR: ultrafast universal RNA-seq aligner." Bioinformatics 29.1 (2013): 15-21.
\end{enumerate}
\subsubsection{Metagenomic classification and abundance estimation (one of the use cases of short-read alignment)}
\begin{enumerate}
    \item Wood, Derrick E., and Steven L. Salzberg. "Kraken: ultrafast metagenomic sequence classification using exact alignments." Genome biology 15.3 (2014): R46.
    \item Lu, Jennifer, et al. "Bracken: estimating species abundance in metagenomics data." PeerJ Computer Science 3 (2017): e104.
    \item Reppell, Mark, and John Novembre. "Using pseudoalignment and base quality to accurately quantify microbial community composition." PLoS computational biology 14.4 (2018): e1006096.
\end{enumerate}
\subsubsection{Long-read alignment}
\begin{enumerate}
    \item Li, Heng. "Minimap2: pairwise alignment for nucleotide sequences." Bioinformatics 34.18 (2018): 3094-3100.
    \item Jain, Chirag, et al. "A fast approximate algorithm for mapping long reads to large reference databases." International Conference on Research in Computational Molecular Biology. Springer, Cham, 2017.
\end{enumerate}
\subsubsection{Fusion Detection → Application of some of the previous tools, methods, and algorithms}
\begin{enumerate}
    \item Dehghannasiri, Roozbeh, et al. "Improved detection of gene fusions by applying statistical methods reveals new oncogenic RNA cancer drivers." bioRxiv (2019): 659078.
    \item Haas, Brian, et al. "STAR-Fusion: fast and accurate fusion transcript detection from RNA-Seq." BioRxiv (2017): 120295.
    \item Liu, Silvia, et al. "Comprehensive evaluation of fusion transcript detection algorithms and a meta-caller to combine top performing methods in paired-end RNA-seq data." Nucleic acids research 44.5 (2015): e47-e47.
\end{enumerate}


